---
output: github_document
title: "Get Rich with 'stocks'"
author: "Dane Van Domelen <br> vandomed@gmail.com"
date: "`r Sys.Date()`"
bibliography: bibliography.bib
nocite: | 
  @quantmod
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

[![Build Status](https://travis-ci.org/vandomed/stocks.svg?branch=master)](https://travis-ci.org/vandomed/stocks)

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  #fig.path = "README-",
  fig.path = "README-figures/",
  warning = FALSE, 
  message = FALSE, 
  out.width = '80%', 
  dpi = 600
)
```

```{r setup, include = FALSE}
options(digits = 3)
```

## 'stocks' package

The **stocks** package has a variety of functions for analyzing investments and investment strategies. I use it for a lot of my articles on [Seeking Alpha](https://seekingalpha.com/author/dane-van-domelen/articles#regular_articles). The package relies heavily on [Yahoo! Finance](https://finance.yahoo.com/) for historical prices and on the **quantmod** package for downloading that data into R.

There are functions for calculating performance metrics, visualizing the performance of funds and multi-fund portfolios, and backtesting trading strategies. The main functions are:

| Function                  |Purpose
|:--------------------------|:-----------------------------------------------------
|`load_prices`              |Download Historical Prices
|`load_gains`               |Download Historical Gains
|`plot_growth`              |Plot Investment Growth
|`calc_metrics`             |Calculate Performance Metrics
|`calc_metrics_overtime`    |Calculate Performance Metrics over Time
|`calc_metrics_2funds`      |Calculate Performance Metrics for Two-Fund Portfolios
|`calc_metrics_3funds`      |Calculate Performance Metrics for Three-Fund Portfolios
|`plot_metrics`             |Plot One Performance Metric (Sorted Bar Plot) or One vs. Another (Scatterplot)
|`plot_metrics_overtime`    |Plot One Performance Metric vs. Time or One vs. Another over Time
|`plot_metrics_2funds`      |Plot One Performance Metric vs. Another for Two-Fund Portfolios
|`plot_metrics_3funds`      |Plot One Performance Metric vs. Another for Three-Fund Portfolios


## Motivating example: A two-fund stocks and bonds portfolio

### Rationale

Stocks and bonds are obviously the primary building blocks for a retirement portfolio, and I think the ETF's SPY and TLT pair together very nicely for a very effective two-fund strategy. Let's look at the performance of these funds separately and together.

### Assess each fund's performance over their mutual lifetimes

We can use `load_gains` to download historical daily gains for SPY and TLT over their mutual lifetimes:

```{r}
library("stocks")
gains <- load_gains(c("SPY", "TLT"), mutual.start = TRUE, to = "2018-12-31")
head(gains)
```

We can call (or pipe into) `calc_metrics` to calculate some performance metrics. `calc_metrics` returns a normal data frame, but I'll call `knitr::kable` to print it as a neat-looking table:

```{r}
metrics <- calc_metrics(gains)
knitr::kable(metrics)
```

We see here that SPY has achieved stronger growth (8.5% vs. 6.3%), but with a much worse max drawdown (55.2% vs. 26.6%). TLT's Sharpe ratio (a measure of risk-adjusted returns) is somewhat higher than SPY's. 

Without getting too far ahead of myself, TLT's positive alpha (0.039%) and negative beta (-0.292) are precisely why it pairs so well with SPY. This isn't unique to TLT; all bond funds should generate alpha (otherwise, don't invest!), and they're often negatively correlated with equities. 

For a visual comparison of the returns and volatility of these two ETF's, we can plot mean vs. SD using `plot_metrics`. 

```{r}
plot_metrics(metrics, mean ~ sd)
```

No surprise, the S&P 500 ETF had more growth, but also higher volatility.

(Side note: You could achieve the same plot by specifying `gains` rather than `metrics`, or by simply specifying the `tickers` input.)


### How reliable is TLT's negative correlation?

Negative correlation works wonders for a two-fund portfolio, so let's look at how consistently TLT achieves negative correlation with SPY, using `calc_metrics_overtime` and `plot_metrics_overtime`. For illustrative purposes, I'll include the full 3-step process: load historical gains, calculate beta over time, and plot beta over time.

```{r}
c("SPY", "TLT") %>%
  load_gains(to = "2018-12-31") %>%
  calc_metrics_overtime("pearson") %>%
  plot_metrics_overtime(pearson ~ .)
```

While the tendency is certainly for negative correlation, there's a lot of variability, and in some years the correlation was actually slightly positive.

As you can see, the default behavior is to calculate the requested metric on a per-year basis. You can also request per-month calculations or rolling windows of a certain width (see `?calc_metrics_overtime`). And the Pearson correlation is just one of many metrics you can plot (see `?calc_metrics` for the full list).

Everyone loves piping these days, but for typical use cases I would actually recommend skipping directly to `plot_metrics_overtime`. If you specify `tickers`, it will download the data it needs on the fly. This code is much shorter and produces the same figure as above:

```{r, eval = FALSE}
plot_metrics_overtime(formula = beta ~ ., tickers = "TLT")
```


### A 50-50 blend

A 50% SPY, 50% TLT portfolio should generate much better risk-adjusted returns than SPY (and perhaps TLT) itself, but a 50% bonds allocation is pretty high so raw returns will probably be lower.

To look at this, we can add a column to `gains` and then call `calc_metrics`, requesting a few particular metrics:

```{r}
gains$`50-50` <- gains$SPY * 0.5 + gains$TLT * 0.5
calc_metrics(gains, c("cagr", "mdd", "sharpe", "sortino")) %>%
  knitr::kable()
```

Indeed, while the 50-50 portfolio achieved slightly lower raw returns than SPY alone, its max drawdown was far better, and its Sharpe and Sortino ratios indicated much better risk-adjusted growth compared to the individual ETF's.


### What's the optimal allocation?

That will likely depend on what metric you want to maximize. In terms of raw growth, roughly 75% SPY is optimal, but the curve is pretty flat--the CAGR is roughly the same from 60-100% SPY.

```{r}
plot_metrics_2funds(gains = gains, 
                    formula = cagr ~ allocation, 
                    tickers = c("SPY", "TLT"))
```

In terms of risk-adjusted growth, the Sharpe ratio curve is somewhat more interesting. The maximum Sharpe ratio occurs around 40% SPY, and the Sharpe ratio gets much worse as you approach 60% SPY and higher.

```{r}
plot_metrics_2funds(gains = gains, 
                    formula = sharpe ~ allocation, 
                    tickers = c("SPY", "TLT"))
```

We can gain additional insight by plotting two metrics against each other, across all possible allocations. A common strategy is to plot the mean vs. standard deviation as a function of the allocation:

```{r}
plot_metrics_2funds(gains = gains, 
                    formula = mean ~ sd, 
                    tickers = c("SPY", "TLT"))
```

This plot yields an interesting finding: starting at 100% TLT, increasing the allocation to SPY simultaneously *reduces volatility* and *increases returns*. In other words, you'd be crazy not to ride the curve up and to the left, adding at least a 30% SPY allocation.

A big caveat is that this is all based on historical data. There's no guarantee that 30% SPY, 70% TLT will have lower volatility *or* greater returns than TLT going forward.


## Three-fund portfolios, plus a stock tip

I won't go into detail about it here, but one of my favorite strategies is UPRO-VBLTX-VWEHX. [UPRO](https://www.proshares.com/funds/upro.html) is a 3x daily S&P 500 ETF and VBLTX and VWEHX are two bonds mutual funds offered by Vanguard.

To visualize the behavior of this 3-fund portfolio, let's plot mean vs. SD as the allocations vary.

```{r}
plot_metrics_3funds(formula = mean ~ sd, 
                    tickers = c("VWEHX", "VBLTX", "UPRO"))
```

By default, SPY is included on the plot, so you can compare metrics to the S&P. Notice that many points on the UPRO-VBLTX-VWEHX surface are higher and to the left of SPY, meaning this strategy has the potential to trump the S&P in terms of risk-reward. 

One idea would be to choose an allocation to match the volatility (i.e, SD) of SPY. At roughly 1% SD, it looks like the higher black curve, which is just UPRO-VBLTX (0% VWEHX), achieves the highest historical returns. Still, I like having VWEHX in there for a second source of alpha.

To see the actual numbers, you can either specify `return = "data"` (or `return = "both"`), or inptu the plot returned by `plot_metrics_3funds` to `plotly::ggplotly`.

I'll close it out with a hot stock tip: buy FANG.

```{r}
plot_growth(tickers = c("FB", "AAPL", "NFLX", "GOOG"), from = "2015-01-01")
```


## Version history

| Version                   |Updates
|:--------------------------|:-----------------------------------------------------
|1.0                        |Original and minor changes
|1.2-1.4                    |Added some new functions
|2.0                        |Switched to **ggplot**, added piping support, simplified functions for calculating metrics


## References
